### 2018-05-25 submission/manuscript.Rmd
make changes inspired by Marcy to intro

**Background.** It is challenging to compare 16S rRNA gene sequencing data across studies and one of the reasons for this is due to error. There are many different places throughout the workflow where error can be introduced into the pipeline. Here, we focus on studying how the number of cycles and high fidelity (HiFi) DNA polymerase introduce error into this workflow because of less investigation into this area. By varying both cycle number and polymerase we assess the changes that occur to the bacterial community in human fecal and mock community samples.

@@

**Results.** When analyzing fecal samples we observed that the range in the number of OTUs detected was not consistent between HiFi DNA polymerases at 35 cycles (Accuprime = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` (min - max) versus Phusion = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`). Additionally, the median number of OTUs vaired by HiFi DNA polymerase used (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001). When analyzing mock community samples the variations in the number of OTUs detected by DNA polymerase was observable as early as 20 cycles (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`). There also was a large range in the number of OTUs by HiFi DNA polymerase at 35 cycles (Accuprime = `r filter(mock_num_otu_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(mock_num_otu_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` versus Phusion = `r filter(mock_num_otu_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(mock_num_otu_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`). Chimera prevalence in mock communities varied by polymerase with differences being most notable at 35 cycles (Kappa = `r round((filter(error_summary, cycles == "35x", taq == "K") %>% pull(chimera_prevalence_median))*100, digits = 2)`% (median) versus Platinum = `r round((filter(error_summary, cycles == "35x", taq == "PL") %>% pull(chimera_prevalence_median))*100, digits = 2)`%) and this variation persisted after chimera removal using VSEARCH. We also observed positive correlations between chimera prevalence and the number of OTUs with Platinum having the highest (R^2 = `r max(pull(r2_table, r2))`) and Kappa having the worse (R^2 = `r min(pull(r2_table, r2))`).

@@

Accounting for bias is critical in order to reach an understanding of bacterial community changes using 16S rRNA gene sequencing results. Bias can change the observed results in a way that is reproducible and standardized. Within this context, our 16S rRNA gene sequencing methods are biased even when these workflows are standardized to increase reproducibility. In order to interpret specific studies within the broader context of the overall field, assessing bias at different parts of the 16S rRNA gene sequencing workflow is critical.

Many parts of the 16S rRNA gene sequencing workflow contribute bias to the results and are studied extensively. A typical 16S rRNA gene sequencing workflow can be divided into preservation, extraction, PCR, and sequencing steps. Overall, biases due to either preservation or extraction tend to be smaller than the overall biological signal being measured [@preservation_Song_2016; @storage_Bassis_2017]. Although the observed biological signal is larger, the contribution of PCR bias to this overall workflow is not well characterized since these studies use the same PCR approach while varying preservation or extraction method.

Identifying the biases in the PCR stage of 16S rRNA gene sequencing is important because a large body of literature shows that there are a variety of steps during PCR that can change the observed results [@Eckert1991; @Burkardt2000]. Many of these sources of biases are made worse as the number of cycles increases [@Wang1996; @Haas2011; @Kebschull2015]. For example, the selective amplification of AT-rich over GC-rich sequences can exaggerate the difference between 16S rRNA genes higher in AT versus those higher in GC [@polz_bias_1998]. Both amplification error and non-specific amplification (e.g. incorrect amplicon size products) can also increase as the number of cycles increases which can drastically change commonly used diversity measures [@BautistadelosSantos2016; @Acinas2005]. Additionally, chimeras can form from an aborted extension step followed by a subsequent priming error and secondary extension and will artificially increase community diversity [@Haas2011]. Although these differences are not necessarily dependent on primer and DNA polymerase used, there are also biases that are.

The intrinsic properties to primers and DNA polymerases chosen can also introduce bias. Primers have variable region dependent binding affinities for different bacteria and depending on the primer pair do not detect specific bacteria (e.g. V1-V3 does not detect *Haemophilus influenzae* and V3-V5 does not detect *Propionibacterium acnes*) [@Meisel2016; @Sze2015 (Table S4)]. Additionally, there are multiple families of DNA polymerases that have their own error rate and proof reading capacity [@polymerase_Ishino_2014]. Interestingly, the influence that these different DNA polymerases can have on the observed 16S rRNA gene sequencing results have not been well studied like some of the other sources of PCR-based bias.

A recent study found clear differences between normal and high fidelity DNA polymerase and that optimization of the PCR protocol could reduce error and chimera generation [@taq_Gohl_2016]. This study also found that regardless of DNA polymerase, the number of Operational Taxonomic Units (OTUs) or taxa generated were not easily reduced using the authors chosen bioinformatic pipeline [@taq_Gohl_2016]. It is natural to extend this line of inquiry and ask if biases in the number of OTUs and chimeras are also dependent on the type of high fidelity DNA polymerase. There is some reason to think that this may be the case since many of these high fidelity DNA polymerases come from different families (e.g. *Taq* belongs to the family A polymerases) and may intrinsically have different error rates that cannot be completely removed with modifications [@polymerase_Ishino_2014].

Although bias introduced due to differences between DNA polymerase and high fidelity DNA polymerase has been investigated for 16S rRNA gene sequencing, the bias caused due to differences between specific high fidelity DNA polymerases has not been. This study will investigate how high fidelity DNA polymerases can bias observed bacterial community results derived from 16S rRNA gene sequencing. We will accomplish this by examining if any of five different types of high fidelity DNA polymerases introduce significant biases into 16S rRNA gene surveys, if this is a cycle dependent phenomenon, and whether they can be removed using a standard bioinformatic pipeline.

>>>>

**Background.** It is challenging to compare 16S rRNA gene sequencing data across studies and one of the reasons for this is due to error. There are many different places throughout the workflow where error can be introduced into the pipeline. Here, we focus on studying how the number of cycles and high fidelity (HiFi) DNA polymerase introduce error by varying cycle number and polymerase used to amplify 16S rRNA genes in human fecal and mock community samples.

@@

**Results.** When analyzing fecal samples we observed that the range in the number of OTUs detected was not consistent between HiFi DNA polymerases at 35 cycles (Accuprime = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` (min - max) versus Phusion = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`). Additionally, the median number of OTUs vaired by HiFi DNA polymerase used (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001). When analyzing mock community samples the variation in the number of OTUs detected by the polymerases was observable as early as 20 cycles (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`). There also was a large range in the number of OTUs amplified by the polymerases at 35 cycles (Accuprime = `r filter(mock_num_otu_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(mock_num_otu_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` versus Phusion = `r filter(mock_num_otu_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(mock_num_otu_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`). Chimera prevalence in mock communities varied by polymerase with differences being most notable at 35 cycles (Kappa = `r round((filter(error_summary, cycles == "35x", taq == "K") %>% pull(chimera_prevalence_median))*100, digits = 2)`% (median) versus Platinum = `r round((filter(error_summary, cycles == "35x", taq == "PL") %>% pull(chimera_prevalence_median))*100, digits = 2)`%) and this variation persisted after chimera removal using VSEARCH. We also observed positive correlations between chimera prevalence and the number of OTUs with Platinum having the highest (R^2 = `r max(pull(r2_table, r2))`) and Kappa having the worst (R^2 = `r min(pull(r2_table, r2))`).

@@

The bacterial community is reported to vary between case and control for a number of diseases []. However, for diseases like obesity, the taxa identified have varied widely depending on the study []. Some of this variation could be due to error introduced during the 16S rRNA gene sequencing workflow. Yet, standardizing a 16S rRNA gene sequencing workflow will ultimately result in a standardized and reproducible bias due to choices made on the methods used for preservation, extraction, PCR, and sequencing. Within this context, all 16S rRNA gene sequencing methods are biased even when these workflows are standardized to increase reproducibility. In order to interpret specific studies within the broader context of the overall field, assessing error at different parts of the 16S rRNA gene sequencing workflow is critical.

A typical 16S rRNA gene sequencing workflow can be divided into preservation, extraction, PCR, and sequencing steps. The preservation and extraction stages of the 16S rRNA gene sequencing workflow have been the most extensively studied []. For preservation and extraction stages of the workflow, it has been consistently found that there are biases based on the kits used, but that these differences are smaller than the overall biological difference measured between samples with different kits [@preservation_Song_2016; @storage_Bassis_2017]. Since these studies use the same PCR approach while varying preservation or extraction method, the contribution of PCR bias to this overall workflow is not well characterized.

There is a large body of literature that shows there are different biases within each of the steps in the PCR stage of 16S rRNA gene sequencing. One source of error is from primers chosen for the 16S rRNA gene sequencing. There are variable region dependent binding affinities for different bacteria, and depending on the primer pair used they do not detect specific bacteria (e.g. V1-V3 does not detect *Haemophilus influenzae* and V3-V5 does not detect *Propionibacterium acnes*) [@Meisel2016; @Sze2015 (Table S4)]. Another source of error is the selective amplification of AT-rich over GC-rich sequences which exaggerate the difference between 16S rRNA genes higher in AT versus those higher in GC content [@polz_bias_1998]. Many of these sources of biases are made worse as the number of cycles increases [@Wang1996; @Haas2011; @Kebschull2015]. For example, both amplification error and non-specific amplification (e.g. incorrect amplicon size products) also can increase as the number of cycles increases. This will increase the number of Operational Taxonomic Units (OTUs) observed and drastically change the values obtained from commonly used diversity measures [@BautistadelosSantos2016; @Acinas2005]. Additionally, as the number of cycles increases more chimeras can form from an aborted extension step that causes a priming error and subsequent secondary extension [@Haas2011]. These chimeras will artificially increase community diversity by increasing the number of OTUs that are observed [@Haas2011]. In addition to these sources of errors, there also are multiple families of DNA polymerases that have their own error rate and proof reading capacity [@polymerase_Ishino_2014]. Interestingly, the influence that these different DNA polymerases can have on the observed 16S rRNA gene sequencing results have not been well studied like some of the other sources of PCR-based bias.

A recent study found differences in the number of OTUS and chimeras between normal and high fidelity DNA polymerases [@taq_Gohl_2016]. The authors could reduce the difference between the two polymerases by optimizing the annealing and extension steps within the PCR protocol [@taq_Gohl_2016]. Yet, within this study there was no comparison made between different high fidelity DNA polymerases. Due to this gap, it is natural to extend this line of inquiry and test if biases in the number of OTUs and chimeras also are dependent on the type of high fidelity DNA polymerase. This study will investigate how high fidelity DNA polymerases can bias observed bacterial community results derived from 16S rRNA gene sequencing. We will accomplish this by examining the number of OTUs, error rate, number of sequences with an error, and chimera prevalence at varying number of cycles in five different high fidelity DNA polymerases

### 2018-05-25 submission/manuscript.tex
make changes inspired by Marcy to intro

\textbf{Background.} It is challenging to compare 16S rRNA gene
sequencing data across studies and one of the reasons for this is due to
error. There are many different places throughout the workflow where
error can be introduced into the pipeline. Here, we focus on studying
how the number of cycles and high fidelity (HiFi) DNA polymerase
introduce error into this workflow because of less investigation into
this area. By varying both cycle number and polymerase we assess the
changes that occur to the bacterial community in human fecal and mock
community samples.

@@

\textbf{Results.} When analyzing fecal samples we observed that the
range in the number of OTUs detected was not consistent between HiFi DNA
polymerases at 35 cycles (Accuprime = 84 - 106 (min - max) versus
Phusion = 84 - 136). Additionally, the median number of OTUs vaired by
HiFi DNA polymerase used (P-value \textless{} 0.0001). When analyzing
mock community samples the variations in the number of OTUs detected by
DNA polymerase was observable as early as 20 cycles (P-value = 0.002).
There also was a large range in the number of OTUs by HiFi DNA
polymerase at 35 cycles (Accuprime = 15 - 20 versus Phusion = 14 - 73).
Chimera prevalence in mock communities varied by polymerase with
differences being most notable at 35 cycles (Kappa = 5.71\% (median)
versus Platinum = 26.62\%) and this variation persisted after chimera
removal using VSEARCH. We also observed positive correlations between
chimera prevalence and the number of OTUs with Platinum having the
highest (R\^{}2 = 0.974) and Kappa having the worse (R\^{}2 = 0.478).

@@

Accounting for bias is critical in order to reach an understanding of
bacterial community changes using 16S rRNA gene sequencing results. Bias
can change the observed results in a way that is reproducible and
standardized. Within this context, our 16S rRNA gene sequencing methods
are biased even when these workflows are standardized to increase
reproducibility. In order to interpret specific studies within the
broader context of the overall field, assessing bias at different parts
of the 16S rRNA gene sequencing workflow is critical.

Many parts of the 16S rRNA gene sequencing workflow contribute bias to
the results and are studied extensively. A typical 16S rRNA gene
sequencing workflow can be divided into preservation, extraction, PCR,
and sequencing steps. Overall, biases due to either preservation or
extraction tend to be smaller than the overall biological signal being
measured (Song et al., 2016; Bassis et al., 2017). Although the observed
biological signal is larger, the contribution of PCR bias to this
overall workflow is not well characterized since these studies use the
same PCR approach while varying preservation or extraction method.

Identifying the biases in the PCR stage of 16S rRNA gene sequencing is
important because a large body of literature shows that there are a
variety of steps during PCR that can change the observed results (Eckert
\& Kunkel, 1991; Burkardt, 2000). Many of these sources of biases are
made worse as the number of cycles increases (Wang \& Wang, 1996; Haas
et al., 2011; Kebschull \& Zador, 2015). For example, the selective
amplification of AT-rich over GC-rich sequences can exaggerate the
difference between 16S rRNA genes higher in AT versus those higher in GC
(Polz \& Cavanaugh, 1998). Both amplification error and non-specific
amplification (e.g.~incorrect amplicon size products) can also increase
as the number of cycles increases which can drastically change commonly
used diversity measures (Acinas et al., 2005; Santos et al., 2016).
Additionally, chimeras can form from an aborted extension step followed
by a subsequent priming error and secondary extension and will
artificially increase community diversity (Haas et al., 2011). Although
these differences are not necessarily dependent on primer and DNA
polymerase used, there are also biases that are.

The intrinsic properties to primers and DNA polymerases chosen can also
introduce bias. Primers have variable region dependent binding
affinities for different bacteria and depending on the primer pair do
not detect specific bacteria (e.g.~V1-V3 does not detect
\emph{Haemophilus influenzae} and V3-V5 does not detect
\emph{Propionibacterium acnes}) (Sze et al., 2015 (Table S4); Meisel et
al., 2016). Additionally, there are multiple families of DNA polymerases
that have their own error rate and proof reading capacity (Ishino \&
Ishino, 2014). Interestingly, the influence that these different DNA
polymerases can have on the observed 16S rRNA gene sequencing results
have not been well studied like some of the other sources of PCR-based
bias.

A recent study found clear differences between normal and high fidelity
DNA polymerase and that optimization of the PCR protocol could reduce
error and chimera generation (Gohl et al., 2016). This study also found
that regardless of DNA polymerase, the number of Operational Taxonomic
Units (OTUs) or taxa generated were not easily reduced using the authors
chosen bioinformatic pipeline (Gohl et al., 2016). It is natural to
extend this line of inquiry and ask if biases in the number of OTUs and
chimeras are also dependent on the type of high fidelity DNA polymerase.
There is some reason to think that this may be the case since many of
these high fidelity DNA polymerases come from different families (e.g.
\emph{Taq} belongs to the family A polymerases) and may intrinsically
have different error rates that cannot be completely removed with
modifications (Ishino \& Ishino, 2014).

Although bias introduced due to differences between DNA polymerase and
high fidelity DNA polymerase has been investigated for 16S rRNA gene
sequencing, the bias caused due to differences between specific high
fidelity DNA polymerases has not been. This study will investigate how
high fidelity DNA polymerases can bias observed bacterial community
results derived from 16S rRNA gene sequencing. We will accomplish this
by examining if any of five different types of high fidelity DNA
polymerases introduce significant biases into 16S rRNA gene surveys, if
this is a cycle dependent phenomenon, and whether they can be removed
using a standard bioinformatic pipeline.

@@

\hypertarget{ref-Burkardt2000}{}
Burkardt H-J. 2000. Standardization and quality control of PCR analyses.
\emph{Clinical Chemistry and Laboratory Medicine} 38. DOI:
\href{https://doi.org/10.1515/cclm.2000.014}{10.1515/cclm.2000.014}.

@@

\hypertarget{ref-Eckert1991}{}
Eckert KA., Kunkel TA. 1991. DNA polymerase fidelity and the polymerase
chain reaction. \emph{Genome Research} 1:17--24. DOI:
\href{https://doi.org/10.1101/gr.1.1.17}{10.1101/gr.1.1.17}.

>>>>

\textbf{Background.} It is challenging to compare 16S rRNA gene
sequencing data across studies and one of the reasons for this is due to
error. There are many different places throughout the workflow where
error can be introduced into the pipeline. Here, we focus on studying
how the number of cycles and high fidelity (HiFi) DNA polymerase
introduce error by varying cycle number and polymerase used to amplify
16S rRNA genes in human fecal and mock community samples.

@@

\textbf{Results.} When analyzing fecal samples we observed that the
range in the number of OTUs detected was not consistent between HiFi DNA
polymerases at 35 cycles (Accuprime = 84 - 106 (min - max) versus
Phusion = 84 - 136). Additionally, the median number of OTUs vaired by
HiFi DNA polymerase used (P-value \textless{} 0.0001). When analyzing
mock community samples the variation in the number of OTUs detected by
the polymerases was observable as early as 20 cycles (P-value = 0.002).
There also was a large range in the number of OTUs amplified by the
polymerases at 35 cycles (Accuprime = 15 - 20 versus Phusion = 14 - 73).
Chimera prevalence in mock communities varied by polymerase with
differences being most notable at 35 cycles (Kappa = 5.71\% (median)
versus Platinum = 26.62\%) and this variation persisted after chimera
removal using VSEARCH. We also observed positive correlations between
chimera prevalence and the number of OTUs with Platinum having the
highest (R\^{}2 = 0.974) and Kappa having the worst (R\^{}2 = 0.478).

@@

The bacterial community is reported to vary between case and control for
a number of diseases {[}{]}. However, for diseases like obesity, the
taxa identified have varied widely depending on the study {[}{]}. Some
of this variation could be due to error introduced during the 16S rRNA
gene sequencing workflow. Yet, standardizing a 16S rRNA gene sequencing
workflow will ultimately result in a standardized and reproducible bias
due to choices made on the methods used for preservation, extraction,
PCR, and sequencing. Within this context, all 16S rRNA gene sequencing
methods are biased even when these workflows are standardized to
increase reproducibility. In order to interpret specific studies within
the broader context of the overall field, assessing error at different
parts of the 16S rRNA gene sequencing workflow is critical.

A typical 16S rRNA gene sequencing workflow can be divided into
preservation, extraction, PCR, and sequencing steps. The preservation
and extraction stages of the 16S rRNA gene sequencing workflow have been
the most extensively studied {[}{]}. For preservation and extraction
stages of the workflow, it has been consistently found that there are
biases based on the kits used, but that these differences are smaller
than the overall biological difference measured between samples with
different kits (Song et al., 2016; Bassis et al., 2017). Since these
studies use the same PCR approach while varying preservation or
extraction method, the contribution of PCR bias to this overall workflow
is not well characterized.

There is a large body of literature that shows there are different
biases within each of the steps in the PCR stage of 16S rRNA gene
sequencing. One source of error is from primers chosen for the 16S rRNA
gene sequencing. There are variable region dependent binding affinities
for different bacteria, and depending on the primer pair used they do
not detect specific bacteria (e.g.~V1-V3 does not detect
\emph{Haemophilus influenzae} and V3-V5 does not detect
\emph{Propionibacterium acnes}) (Sze et al., 2015 (Table S4); Meisel et
al., 2016). Another source of error is the selective amplification of
AT-rich over GC-rich sequences which exaggerate the difference between
16S rRNA genes higher in AT versus those higher in GC content (Polz \&
Cavanaugh, 1998). Many of these sources of biases are made worse as the
number of cycles increases (Wang \& Wang, 1996; Haas et al., 2011;
Kebschull \& Zador, 2015). For example, both amplification error and
non-specific amplification (e.g.~incorrect amplicon size products) also
can increase as the number of cycles increases. This will increase the
number of Operational Taxonomic Units (OTUs) observed and drastically
change the values obtained from commonly used diversity measures (Acinas
et al., 2005; Santos et al., 2016). Additionally, as the number of
cycles increases more chimeras can form from an aborted extension step
that causes a priming error and subsequent secondary extension (Haas et
al., 2011). These chimeras will artificially increase community
diversity by increasing the number of OTUs that are observed (Haas et
al., 2011). In addition to these sources of errors, there also are
multiple families of DNA polymerases that have their own error rate and
proof reading capacity (Ishino \& Ishino, 2014). Interestingly, the
influence that these different DNA polymerases can have on the observed
16S rRNA gene sequencing results have not been well studied like some of
the other sources of PCR-based bias.

A recent study found differences in the number of OTUS and chimeras
between normal and high fidelity DNA polymerases (Gohl et al., 2016).
The authors could reduce the difference between the two polymerases by
optimizing the annealing and extension steps within the PCR protocol
(Gohl et al., 2016). Yet, within this study there was no comparison made
between different high fidelity DNA polymerases. Due to this gap, it is
natural to extend this line of inquiry and test if biases in the number
of OTUs and chimeras also are dependent on the type of high fidelity DNA
polymerase. This study will investigate how high fidelity DNA
polymerases can bias observed bacterial community results derived from
16S rRNA gene sequencing. We will accomplish this by examining the
number of OTUs, error rate, number of sequences with an error, and
chimera prevalence at varying number of cycles in five different high
fidelity DNA polymerases

@@

@@
